---
title: "Kendall Model"
author: "Kendall Fitzgerald"
date: "2025-04-04"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Invertebrates to include in analysis: purple urchin, sunflower sea star, northern kelp crab

## Setup
```{r}
#load necessary packages
library(here)
library(tidyverse)
library(naniar)
library(DHARMa)
library(gtsummary)
library(broom)
library(MASS)

#load dataset
inverts_kelp <- read_csv("PISCO_kelpforest_swath.1.2.csv")

```

```{r}
#tidy data
inverts_kelp <- inverts_kelp %>%
  filter(campus == "UCSC",
         classcode %in% c("PUGPRO", "STRPURAD", "PYCHEL", "MACPYRAD")) %>%
  #only include relevant columns for further analysis
  select(campus, survey_year, site, zone, classcode, count, size, disease, depth)

#aggregate counts by site/year/species
inverts_summarized <- inverts_kelp %>%
  group_by(campus,site, survey_year, classcode, depth) %>%
  summarise(
    total_count = sum(count, na.rm = TRUE),
    zone = first(zone),
    .groups = "drop"
  )

#pivot to wide format
inverts_wide <- inverts_summarized %>%
  pivot_wider(
    names_from = classcode,
    values_from = total_count,
    values_fill = list(total_count = 0)
  )

 
```

## Step 1. Define Research Question
How does the presence of different invertebrates correlate with kelp abundance?

Null Hypothesis: Invertebrate presence does not have any correlation with kelp abundance.

Alternative Hypothesis: Invertebrate presence does have a correlation with kelp abundance.

## Step 2. Examine data and possible correlations
```{r}
#create and examine figures of raw values

#histograms of species count

#histogram of giant kelp
kelp_fig <- ggplot(inverts_wide, aes(y = MACPYRAD,
                                      x = survey_year)) +
  geom_bar(stat = "identity") +
   labs(y = "Giant Kelp",
       x = "Survey Year") +
  theme_bw()

kelp_fig

#histogram of Northern kelp crab
crab_fig <- ggplot(inverts_wide, aes(y = PUGPRO,
                                      x = survey_year)) +
  geom_bar(stat = "identity") +
  labs(y = "Northern Kelp Crab",
       x = "Survey Year") +
  theme_bw()

crab_fig

#histogram of Sunflower Sea Star
seastar_fig <- ggplot(inverts_wide, aes(y = PYCHEL,
                                      x = survey_year)) +
  geom_bar(stat = "identity") +
     labs(y = "Sunflower Sea Star",
       x = "Survey Year") +
  theme_bw()

seastar_fig

#histogram of purple urchin 
urchin_fig <- ggplot(inverts_wide, aes(y = STRPURAD,
                                      x = survey_year)) +
  geom_bar(stat = "identity") +
     labs(y = "Purple Urchin",
       x = "Survey Year") +
  theme_bw()

urchin_fig



```

