---
title: "Kendall Model"
author: "Kendall Fitzgerald"
date: "2025-04-04"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Invertebrates to include in analysis: purple urchin, sunflower sea star, northern kelp crab

## Setup
```{r warning=FALSE}
#load necessary packages
library(here)
library(tidyverse)
library(dplyr)
library(naniar)
library(gtsummary)
library(lme4)
library(MASS)

#load dataset
inverts_kelp <- read_csv("PISCO_kelpforest_swath.1.2.csv")

```

```{r message=FALSE}
#tidy data, the addition of recruit and juvenile classcodes does not make a difference so will move forward with only adult species 
inverts_kelp <- inverts_kelp %>%
  filter(campus == "UCSC",
         classcode %in% c("PUGPRO", "STRPURAD", "PYCHEL", "MACPYRAD")) %>%
  #only include relevant columns for further analysis
  dplyr::select(campus, survey_year, site, zone, classcode, count, size, disease, depth)

#further tidy data based on relevant rows and sum count data 
inverts_summarized <- inverts_kelp %>%
  group_by(campus, site, survey_year, classcode, depth) %>%
  summarise(
    total_count = sum(count, na.rm = TRUE),
    zone = first(zone),
    .groups = "drop"
  )

#group by species (classcode), year (survey_year), and zone
inverts_wide_grouped <- inverts_summarized %>%
  group_by(classcode, survey_year, zone) %>%
  summarise(across(c(total_count), sum, na.rm = TRUE), .groups = "drop")

#pivot to wide format to separate out species into separate rows
inverts_wide <- inverts_wide_grouped %>%
  pivot_wider(names_from = classcode, values_from = total_count, values_fill = list(total_count = 0))

```

## Step 1. Define Research Question
How does the presence of different invertebrates and habitat zone correlate with kelp abundance?

Null Hypothesis: Invertebrate presence and habitat zone do not have any correlation with kelp abundance.

Alternative Hypothesis: Invertebrate presence and habitat zone do have a correlation with kelp abundance.

## Step 2. Examine data and possible correlations
```{r warning=FALSE}
#create and examine figures of raw values

#histograms of species count

#histogram of giant kelp
kelp_fig <- ggplot(inverts_wide, aes(x = MACPYRAD)) +
  geom_histogram() +
   labs(x = "Giant Kelp",
        title = "Giant Kelp Abundance in Northern CA Kelp Forests") +
  theme_bw()

kelp_fig

#histogram of Northern kelp crab
crab_fig <- ggplot(inverts_wide, aes(x = PUGPRO)) +
  geom_histogram() +
   labs(x = "Northern Kelp Crab",
        title = "Northern Kelp Crab Abundance in Northern CA Kelp Forests")
crab_fig

#histogram of Sunflower Sea Star
seastar_fig <- ggplot(inverts_wide, aes(x = PYCHEL)) +
  geom_histogram() +
   labs(x = "Sunflower Sea Star",
        title = "Sunflower Sea Star Abundance in Northern CA Kelp Forests") +
  theme_bw()

seastar_fig

#histogram of purple urchin 
urchin_fig <- ggplot(inverts_wide, aes(x = STRPURAD)) +
  geom_histogram() +
   labs(x = "Purple Urchin",
        title = "Purple Urchin Abundance in Northern CA Kelp Forests") +
  theme_bw()

urchin_fig

#bar graph of zone distribution (categorical variable)
zone_fig <- ggplot(inverts_kelp, aes(x = zone,
                                     y = count,
                                     fill = zone)) +
  geom_bar(stat = "identity") +
     labs(y = "Count",
       x = "Zone",
       title = "Transect Survey Zone Distribution") +
  theme_bw()

zone_fig

```

### Relationship figures of species through time 
```{r warning=FALSE, fig.cap = "This figure shows the abundance of species used in the model over time. The black dotted line sitting at the year 2013 indicates the first large marine heatwave during that time period. This heatwave significantly contributed to the alteration of the kelp forest ecosystem."}
#first tidy data so that it can be more easily understood by ggplot; summarize the counts by species and year (not separated by zone)
species_counts <- inverts_summarized %>%
  group_by(classcode, survey_year) %>%
  summarise(total_count = sum(total_count, na.rm = TRUE), .groups = "drop")

#create labels for graph
labels <- c("MACPYRAD"="Giant Kelp", 
                              "PUGPRO"="Northern Kelp Crab",
                              "PYCHEL"="Southern Sea Star", 
                              "STRPURAD" = "Purple Urchin")

#create facet-wrapped figure to show species counts by year
species_separate <- ggplot(species_counts, aes(x = survey_year, y = total_count)) +
  geom_point(aes(color = classcode), size = 2, alpha = 0.7) +
  geom_smooth(aes(color = classcode), method = "loess", se = FALSE) +
  geom_vline(xintercept = 2013, linetype = "longdash", color = "black")  +
  facet_wrap(~ classcode, scales = "free_y", labeller = labeller(classcode = labels))+
  labs(title = "Distribution of Kelp Forest Species Abundance Used in Model Over Time",
       x = "Survey Year",
       y = "Total Count") +
  theme_bw() +
  theme(strip.text = element_text(face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5, face = "bold", size = 12),
        legend.position = "none")

species_separate

```


### Performing correlation tests between independent variables
```{r, warning = FALSE}
#make dataframe with continuous predictor variables only
predictors <- inverts_wide %>%
  dplyr::select(PUGPRO, STRPURAD, PYCHEL)

#calculate the correlation matrix
cor_matrix <- cor(predictors, use = "complete.obs")

# View the correlation matrix
cor_matrix

```

Because correlation does not exceed 0.6 between all variables, the test showed little to moderate correlation between the continuous predictor variables. Since this is the case, we can move forward with step 3. 

## Step 3 - Fit regression model
```{r, warning = FALSE}
#check the mean and variance of the count variable
mean(inverts_wide$MACPYRAD)
var(inverts_wide$MACPYRAD)

#because the variance is significantly higher than the mean, we cannot use Poisson regression so we will move forward with negative binomial regression due to overdispersion

#fit a negative binomial regression model
kelp_glm_nb <- glm.nb(MACPYRAD ~ PUGPRO + STRPURAD + PYCHEL + zone,
                       data = inverts_wide)

```


## Step 4 - Evaluate Model Diagnostics
```{r, warning = FALSE}
#Examine overall model output
summary(kelp_glm_nb)

#plot model output residuals  
plot(kelp_glm_nb)

#take out relevant outliers based on residual plots
inverts_final <- inverts_wide[-c(26, 29, 63), ]

#refit model without outliers
kelp_glm_refit <- glm.nb(MACPYRAD ~ PUGPRO + STRPURAD + PYCHEL + zone,
                       data = inverts_final)

#re-examine model output
summary(kelp_glm_refit)

#re-examine model residuals
plot(kelp_glm_refit)

#perform Akaike Information Criterion test to compare the two models
AIC(kelp_glm_nb, kelp_glm_refit)

```

The results of the Akaike Information Criterion (AIC) test showed that the re-fit model had a lower AIC score and, therefore, was a better fit of the data.

## Step 5 - Interpret the model and communicate the results
```{r eval=FALSE, include=FALSE}


```

The Negative Binomial regression model suggests that the abundance of Giant Kelp in study areas of Northern California kelp forests is significantly influenced by the presence of Sunflower Sea Stars (B = 0.004, p = 3.62e-05) and Purple Urchins (B = -9.691e-06, p = 0.0003). Northern kelp crabs (B = -0.0156, p = 0.42), Mid Zone (B = 0.176, p = 0.266), and Outer Zone (B = -0.295, p = 0.0878) by contrast are not significant predictors for Giant Kelp abundance. The significant negative relationship between Giant Kelp and Purple Urchins shows that as urchin populations increase, kelp abundance decreases which makes logical sense as urchins predate on kelp. When left unchecked, they can turn entire kelp forests into "Urchin Barrens" with little to no kelp population. The strong positive relationship between Sunflower Sea Star abundance and kelp abundance also supports this narrative, as Sea Stars primarily predate on urchins and keep their population in check, which in turn helps the kelp population.