---
title: "Kendall Model"
author: "Kendall Fitzgerald"
date: "2025-04-04"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Invertebrates to include in analysis: purple urchin, sunflower sea star, northern kelp crab

## Setup
```{r}
#load necessary packages
library(here)
library(tidyverse)
library(naniar)
library(DHARMa)
library(gtsummary)
library(broom)

#load dataset
inverts_kelp <- read_csv("PISCO_kelpforest_swath.1.2.csv")

```

```{r}
#tidy data
inverts_kelp <- inverts_kelp %>%
  filter(campus == "UCSC",
         classcode %in% c("PUGPRO", "STRPURAD", "PYCHEL", "MACPYRAD")) %>%
  #only include relevant columns for further analysis
  select(campus, survey_year, site, zone, classcode, count, size, disease, depth)
 
```

## Step 1. Define Research Question
How does the presence of different invertebrates correlate with kelp abundance?

Null Hypothesis: Invertebrate presence does not have any correlation with kelp abundance.

Alternative Hypothesis: Invertebrate presence does not have a correlation with kelp abundance.

## Step 2. Examine data and possible correlations
```{r}
#create and examine figures of raw values

#separate out species for this step
kelp <- inverts_kelp %>%
  filter(classcode == "MACPYRAD")

nk_crab <- inverts_kelp %>%
  filter(classcode == "PUGPRO")

sun_star <- inverts_kelp %>%
  filter(classcode == "PYCHEL")

purp_urchin <- inverts_kelp %>%
  filter(classcode == "STRPURAD")

#histograms of species count

#labels for species
labels <- c("MACPYRJUV"="Giant Kelp", 
            "PUGPRO"="Northern Kelp Crab",
            "STRPURAD"="Purple Urchin", 
            "PYCHEL" = "Sunflower Sea Star")

#histogram of giant kelp
kelp_fig1 <- ggplot(kelp, aes(x = survey_year, 
                                      y = count)) +
  geom_bar(stat = "identity") +
  theme_bw()
  
kelp_fig1

#histogram of Northern kelp crab
crab_fig1 <- ggplot(nk_crab, aes(x = survey_year, 
                                      y = count)) +
  geom_bar(stat = "identity") +
  theme_bw()

crab_fig1

#histogram of Sunflower Sea Star
seastar_fig1 <- ggplot(sun_star, aes(x = survey_year, 
                                      y = count)) +
  geom_bar(stat = "identity") +
  theme_bw()

seastar_fig1

#histogram of purple urchin 
urchin_fig1 <- ggplot(purp_urchin, aes(x = survey_year, 
                                      y = count)) +
  geom_bar(stat = "identity") +
  theme_bw()

urchin_fig1



```

