---
title: "Mason_Model"
output: html_document
---

## Setup
```{r}
# Load Packages
library(here)
library(ggplot2)
library(tidyverse)
library(naniar)
library(gtsummary)
library(broom)
library(DHARMa)
library(lme4)
library(dplyr)

# Load in data
kelp_data <- read_csv("PISCO_kelpforest_swath.1.2.csv")

fish_data <- read_csv("pisco_fish.csv")

# custom_theme
custom_theme <- function() {
  theme_minimal() +  # Use minimal theme for a clean look
  theme(
    plot.title = element_text(size = 16, face = "bold"),  # Customize title
    plot.subtitle = element_text(size = 12, face = "italic"),  # Subtitle style
    axis.title.x = element_text(size = 14),  # X-axis title style
    axis.title.y = element_text(size = 14),  # Y-axis title style
    axis.text.x = element_text(angle = 45, hjust = 1, size = 12),  # Rotate x-axis labels
    axis.text.y = element_text(size = 12),  # Y-axis text style
    strip.text = element_text(size = 12, face = "bold"),  # Facet labels customization
    plot.caption = element_text(size = 10, face = "italic"),  # Caption style
    legend.position = "top"  # Position the legend at the top for clarity
  )
}
```

# Step 1: Define the Research Question

# Research Question: What environmental and biological factors (e.g., depth, size, species, site) are associated with different severities of disease (or presence of disease?) in invertebrates in UCSC kelp forests?

# Hypotheses:
# Null Hypothesis (H₀): None of the predictors (depth, size, species, etc.) are significantly associated with disease severity (or presence?).
# Alternative Hypothesis (H₁): At least one of the predictors significantly affects disease severity (or presence?).

### Tidy Data
```{r}
kelp_disease_data <- kelp_data %>% 
  # filter and change disease data for data analysis
  filter(disease %in% c("SEVERE", "MILD", "HEALTHY", "YES", "WASTING", "BLACK SPOT")) %>%
  mutate(disease_severity = factor(
    case_when(
      disease == "HEALTHY" ~ "Healthy",  # Healthy
      disease == "MILD" ~ "Mild",        # Mild
      disease == "YES" ~ "Mild",         # Yes treated as Mild
      disease == "WASTING" ~ "Severe",   # Wasting treated as Severe
      disease == "BLACK SPOT" ~ "Severe",# Black spot treated as Severe
      disease == "SEVERE" ~ "Severe"     # Severe
    ),
   levels = c("Healthy", "Mild", "Severe")
  )) %>% 
  # Make disease data binary
 mutate(disease_severity_binary = ifelse(disease_severity == "Healthy", 0, 1)) %>% 
  filter(!is.na(disease_severity)) %>%
  # Filter to only UCSC Campus
  filter(campus == "UCSC") %>%
  # Select only the species most affected by disease, change to their common names
  mutate(
    invertebrates = case_when(
      classcode == "PATMIN" ~ NA_character_,
      classcode == "PISGIG" ~ "Giant Spined Star",
      classcode == "STRPURAD" ~ "Purple Urchin Adult",
      classcode == "DERIMB" ~ NA_character_,
      classcode == "MESFRAAD" ~ "Red Urchin Adult",
      classcode == "PISOCH" ~ NA_character_,
      classcode == "LEPHEXAD" ~ NA_character_,
      classcode == "STRPURREC" ~ NA_character_,
      classcode == "MESFRAREC" ~ NA_character_,
      classcode == "MEDAEQ" ~ NA_character_,
      classcode == "PISBRE" ~ NA_character_,
      classcode == "PISSPP" ~ NA_character_,
      TRUE ~ NA_character_
    )) %>%
  filter(!is.na(invertebrates)) %>%
  filter(!is.na(size) & size > 0) %>% 
  # Change character variables to factors for analysis
  mutate(
    month = factor(month),
    year = factor(year),
    site = factor(site),
    species = factor(invertebrates),
    transect = factor(transect, levels = c("1", "2")),
    zone = factor(zone, levels = c("INNER",
                                   "MID",
                                   "OUTER"))) %>% 
  dplyr::select(-observer, -notes, -site_name_old, -campus, -method,
                -survey_year, -disease, -day, -classcode, -invertebrates) %>% 
  drop_na()
```

### 2. Examine Data and Possible Correlations

# Check for missing values
```{r}
summary(kelp_disease_data)
sapply(kelp_disease_data, function(x) sum(is.na(x)))
```
# Visualize distributions
```{r}
# Calculate optimal bins using the Rice rule
kelp_disease_data <- kelp_disease_data %>%
  mutate(optimal_bins = round(2 * (n() / 3)^(1/3)))


# Visualize distribution of disease (binary outcome)
disease_counts <- kelp_disease_data %>%
  group_by(disease_severity, species) %>%
  summarize(total_count = sum(count), .groups = "drop") %>%
  ggplot(aes(x = disease_severity, y = total_count, fill = species)) +
  geom_bar(stat = "identity", color = "black") +
  scale_fill_manual(
    values = c(
      "Red Urchin Adult" = "red",
      "Purple Urchin Adult" = "purple",
      "Giant Spined Star" = "green"
    ),
    name = "Species"
  ) +
  labs(
    title = "Total Number of Invertebrates by Disease Severity and Species",
    x = "Disease Severity (Health Status)",
    y = "Log Transformed Total Invertebrate Count",
    subtitle = "Distribution of disease severity by species",
    caption = "Data Source: Kelp Disease Dataset"
  ) +
  facet_wrap(~ species) +
  scale_y_continuous(trans = 'log10', labels = scales::comma) +
  custom_theme()




# Visualize distribution of depth
kelp_disease_data_summary <- kelp_disease_data %>%
  group_by(depth, species) %>%
  summarize(total_count = sum(count), .groups = "drop")

# Visualize the summed count of species by depth
depth_counts <- ggplot(kelp_disease_data_summary, aes(x = depth, y = total_count, fill = species)) +
  geom_bar(stat = "identity", color = "black", alpha = 0.7) +
  scale_fill_manual(
    values = c(
      "Red Urchin Adult" = "red",
      "Purple Urchin Adult" = "purple",
      "Giant Spined Star" = "green"
    ),
    name = "Species"
  ) +
  labs(
    title = "Total Number of Invertebrates by Depth",
    x = "Depth (meters)",
    y = "Log Transformed Total Invertebrate Count",
    subtitle = "Total number of invertebrates by depth and species",
    caption = "Data Source: Kelp Disease Dataset"
  ) +
  facet_wrap(~ species) +
  scale_y_continuous(trans = 'log10', labels = scales::comma) +
  custom_theme()







# Visualize distribution of size
size_counts <- kelp_disease_data %>%
  group_by(size, species) %>%
  summarize(total_count = sum(count), .groups = "drop") %>%
  ggplot(aes(x = size, y = total_count, fill = species)) +
  geom_bar(stat = "identity", color = "black", alpha = 0.7) +
  scale_fill_manual(
    values = c(
      "Red Urchin Adult" = "red",
      "Purple Urchin Adult" = "purple",
      "Giant Spined Star" = "green"
    ),
    name = "Species"
  ) +
  labs(
    title = "Total Number of Invertebrates by Size",
    x = "Size (cm)",
    y = "Log Transformed Total Count of Invertebrates",
    subtitle = "Total number of invertebrates by size and species",
    caption = "Data Source: Kelp Disease Dataset"
  ) +
  facet_wrap(~ species) +
  scale_y_continuous(trans = 'log10', labels = scales::comma) +
  custom_theme()





# Visualize distribution of species (categorical)
invert_counts <- kelp_disease_data %>%
  group_by(species) %>%
  summarize(total_count = sum(count), .groups = "drop") %>%
  ggplot(aes(x = species, y = total_count, fill = species)) +
  geom_bar(stat = "identity", color = "black", alpha = 0.7) +
  scale_fill_manual(
    values = c(
      "Red Urchin Adult" = "red",
      "Purple Urchin Adult" = "purple",
      "Giant Spined Star" = "green"
    ),
    name = "Species"
  ) +
  labs(
    title = "Total Number of Each Invertebrate Species",
    x = "Species",
    y = "Log Transformed Total Count",
    subtitle = "Total number of individuals per species",
    caption = "Data Source: Kelp Disease Dataset"
  ) +
  scale_y_continuous(trans = 'log10', labels = scales::comma) + 
  custom_theme() 





# Visualize distribution of zones (categorical)
zone_counts <- kelp_disease_data %>%
  group_by(zone, species) %>%
  summarize(total_count = sum(count), .groups = "drop") %>%
  ggplot(aes(x = zone, y = total_count, fill = species)) +
  geom_bar(stat = "identity", color = "black", alpha = 0.7) +
  scale_fill_manual(
    values = c(
      "Red Urchin Adult" = "red",
      "Purple Urchin Adult" = "purple",
      "Giant Spined Star" = "green"
    ),
    name = "Species"
  ) +
  labs(
    title = "Total Number of Invertebrates by Zone",
    x = "Zone",
    y = "Log Transformed Total Invertebrate Count",
    subtitle = "Total number of individuals per species in each zone",
    caption = "Data Source: Kelp Disease Dataset"
  ) +
  facet_wrap(~ species) +
  scale_y_continuous(trans = 'log10', labels = scales::comma) +
  custom_theme() 



# View Plots
invert_counts
size_counts
depth_counts
disease_counts
zone_counts

```
# Check for correlations between continuous predictors (depth, size)
```{r}
kelp_disease_data_grouped <- kelp_disease_data %>%
  group_by(depth, size, species, month, zone, site, transect, disease_severity_binary) %>%
  summarize(total_count = sum(count), .groups = "drop")

cor(kelp_disease_data_grouped %>%
       dplyr::select(depth, size))
```

### 3. Fit Regression Model
```{r}
model_binary <- glmer(
  disease_severity_binary ~ depth + size + species + month + zone + 
    (1 | site) + (1 | transect),
  data = kelp_disease_data_grouped,
  family = binomial
)

# Summary of the model
summary(model_binary)

# transect not having much effect


# Remove Outliers
# Calculate Cook's Distance
cooks_distance <- cooks.distance(model_binary)

# Filter out observations with high Cook's distance (threshold = 4/n)
kelp_disease_data_cooks <- kelp_disease_data[which(cooks_distance < (4 / nrow(kelp_disease_data))), ]


# Without Transect and outliers
model_binary2 <- glmer(
  disease_severity_binary ~ depth + size + species + month + zone + 
    (1 | site),
  data = kelp_disease_data_cooks,
  family = binomial
)

# Summary of the model
summary(model_binary2)
tbl_regression(model_binary2)



```

### 4. Evaluate Model Diagnostics
```{r}
# Extract residuals and fitted values
kelp_disease_data_cooks$residuals <- resid(model_binary2)
kelp_disease_data_cooks$fitted_values <- fitted(model_binary2)

# DHARMA residual plot for glmer
# we did inspect residuals and removed outliers

# Residuals vs Fitted Plot
plot(kelp_disease_data_cooks$fitted_values, kelp_disease_data_cooks$residuals, 
     xlab = "Fitted Values", ylab = "Residuals",
     main = "Residuals vs Fitted Values")
abline(h = 0, col = "red")

# DHARMA
# Simulate residuals
simulation_output <- simulateResiduals(fittedModel = model_binary2)

# Plot diagnostic plots
plot(simulation_output)

# Normal Q-Q Plot
qqnorm(kelp_disease_data_cooks$residuals)
qqline(kelp_disease_data_cooks$residuals, col = "red")

# Scale-Location Plot (Spread-Location)
sqrt_residuals <- sqrt(abs(kelp_disease_data_cooks$residuals))
plot(kelp_disease_data_cooks$fitted_values, sqrt_residuals, 
     xlab = "Fitted Values", ylab = "Sqrt(|Residuals|)",
     main = "Scale-Location Plot")
abline(h = 0, col = "red")

# Residuals vs Leverage Plot (using hat-values)
hat_values <- hatvalues(model_binary2)
plot(hat_values, kelp_disease_data_cooks$residuals,
     xlab = "Leverage (Hat-Values)", ylab = "Residuals",
     main = "Residuals vs Leverage")
abline(h = 0, col = "red")

testOutliers(simulation_output)
```


# Visualize Model Output
```{r}
# Get the fixed effects
fixed_effects <- summary(model_binary2)$coefficients
fixed_effects_df <- data.frame(
  term = rownames(fixed_effects),
  estimate = fixed_effects[, "Estimate"],
  std_error = fixed_effects[, "Std. Error"],
  z_value = fixed_effects[, "z value"],
  p_value = fixed_effects[, "Pr(>|z|)"]
)


# Add confidence intervals
fixed_effects_df$conf_low <- fixed_effects_df$estimate - 1.96 * fixed_effects_df$std_error
fixed_effects_df$conf_high <- fixed_effects_df$estimate + 1.96 * fixed_effects_df$std_error

# Modify predictor names
fixed_effects_df$term <- gsub("zoneOUTER", "Outer Zone", fixed_effects_df$term)
fixed_effects_df$term <- gsub("month11", "November", fixed_effects_df$term)
fixed_effects_df$term <- gsub("zoneMID", "Mid Zone", fixed_effects_df$term)
fixed_effects_df$term <- gsub("month9", "September", fixed_effects_df$term)
fixed_effects_df$term <- gsub("size", "Size (cm)", fixed_effects_df$term)
fixed_effects_df$term <- gsub("month10", "October", fixed_effects_df$term)
fixed_effects_df$term <- gsub("depth", "Depth (m)", fixed_effects_df$term)
fixed_effects_df$term <- gsub("month7", "July", fixed_effects_df$term)
fixed_effects_df$term <- gsub("month8", "August", fixed_effects_df$term)
fixed_effects_df$term <- gsub("speciesRed Urchin Adult", "Red Urchin", fixed_effects_df$term)
fixed_effects_df$term <- gsub("speciesPurple Urchin Adult", "Purple Urchin", fixed_effects_df$term)

# Define the order of the predictors
predictor_order <- c("Depth (m)", "Size (cm)", "Purple Urchin", "Red Urchin", 
                     "July", "August", "September", "October", "November", 
                     "Mid Zone", "Outer Zone")

# Reorder the 'term' factor based on the custom order
fixed_effects_df$term <- factor(fixed_effects_df$term, levels = predictor_order)

# Remove rows with NA in the term column before plotting
fixed_effects_df <- fixed_effects_df[!is.na(fixed_effects_df$term), ]

# Enhanced plot with reordered predictors
ggplot(fixed_effects_df, aes(x = term, y = estimate)) +
  geom_point(color = "#2C3E50", size = 4, shape = 16) +
  geom_errorbar(aes(ymin = conf_low, ymax = conf_high), 
                width = 0.2, color = "#2980B9", size = 1) +
  coord_flip() +
  theme_minimal(base_size = 15) + 
  theme(
    axis.text.x = element_text(color = "#7F8C8D", size = 12),
    axis.text.y = element_text(color = "#7F8C8D", size = 12),
    axis.title.x = element_text(color = "#34495E", size = 14, face = "bold"),
    axis.title.y = element_text(color = "#34495E", size = 14, face = "bold"),
    plot.title = element_text(size = 16, face = "bold", color = "#2C3E50", hjust = 0.5)
  ) +
  labs(x = "Predictor", y = "Estimate", title = "Fixed Effects with 95% Confidence Intervals")
```
